# Top-down assembly

To begin our explorations of assembly, we consider the case in which all species are introduced in the habitat at the same time, and at arbitrary initial densities.

Here we investigate the simplest case in which dynamics are given by the GLV model, $A$ is symmetric and stable (and therefore Lyapunov Diagonally stable) matrix with random coefficients, and $r$ is a vector of growth rates with random coefficients. For simplicity, we take the off-diagonal elements of $A$ and the growth rates from a normal distribution centered at zero (the same would be found for any distribution symmetric about zero). This case was studied by @servan2018coexistence.

First, we are going to consider a trivial case, then simulate a more complex one, and finally outline the proof in @servan2018coexistence.

## A trivial case

Suppose that the growth rates are drawn from a normal distribution centered at zero, and that the matrix $A$ is diagonal and stable (i.e., species do not interact with each other; each species is self regulating). As such, a species will persist if and only if $r_i > 0$. If we draw the growth rates from any distribution centered at zero, then the probability of having a positive growth rate is $1/2$. Therefore, the number of coexisting species will follow the binomial distribution with parameters $n$ and $1/2$.

## Simulating final composition

Now, let's simulate cases in which species do interact with each other. First, we load the functions we've written before:

```{r, glvfunctionsrz, message=FALSE, warning=FALSE}
# this file contains functions to integrate GLV dynamics
# and plot the results
source("code/general_code.R")
```

And then write functions to give us a (barely) stable, symmetric matrix $A$, and random growth rates:

```{r builddstable}
# function to build symmetric, Lyapunov Diagonally-stable matrix
build_LDstable <- function(n){
  A <- matrix(0, n, n)
  A[upper.tri(A)] <- rnorm(n * (n - 1) / 2)
  # make symmetric
  A <- A + t(A)
  # now find the largest eigenvalue
  l1A <- max(eigen(A, only.values = TRUE, symmetric = TRUE)$values)
  if (l1A > 0){
    # set the diagonal to make it stable
    diag(A) <- diag(A) - l1A - 0.01
  }
  return(A)
}

# function to get random growth rates
build_randomr <- function(n){
  return(rnorm(n))
}
```

Now, we build a random system with seven species, and integrate the dynamics:

```{r rndzooex1}
set.seed(5) # for reproducibility
n <- 7
A <- build_LDstable(n)
r <- build_randomr(n)
x0 <- runif(n)
out <- GLV_dynamics(z0 = x0, A = A, r = r, maxtime = 25, bytime = 0.1)
show(plot_dynamics(out))
knitr::kable(
out$ts %>% 
  filter(out$ts$time == max(out$ts$time)) %>% 
  select(-time) )
```

As you can see, in this case, two species (3 and 5) go extinct, while the other ones reach a feasible equilibrium. Let's try to start the system with species 3 and 5 at high abundance, to show that the equilibrium is indeed globally stable:

```{r rndzooex2}
x0[3] <- 10
x0[5] <- 10
out <- GLV_dynamics(z0 = x0, A = A, r = r, maxtime = 25, bytime = 0.1)
show(plot_dynamics(out))
knitr::kable(
out$ts %>% 
  filter(out$ts$time == max(out$ts$time)) %>% 
  select(-time) )
```

As you can see, the system still goes to the same equilibrium, with 3 and 5 extinct. Turns out, for this type of system, one does not even need to integrate dynamics: the Lemke–Howson algorithm can be adapted to solve the problem efficiently [@servan2018coexistence]. I have coded up the algorithm already, and you can load the function `get_final_composition(A, r)` by typing:

```{r rndzooex3}
source("code/L-H.R")
get_final_composition(A, r) # use LH instead of integrating dynamics
```

## A random zoo

In the case above, 5 species persisted. We can ask how many species will coexist in general under this parametrization. A good metaphor is that of the random zoo: take a large zoo, open all the cages, and return after fifty years. How many species will you find?

To simulate the random zoo, we take random matrices and random growth rates with $n$ species, and tally the number of coexisting species over several simulations:

```{r randomzoo, warning=FALSE}
set.seed(1)
n <- 5
nsim <- 2000
results <- tibble(simulation = 1:nsim, 
                      ncoexisting = rep(NA, nsim))
for (i in 1:nsim){
  # build the matrix and vector
  A <- build_LDstable(n)
  r <- build_randomr(n)
  xstar <- get_final_composition(A, r)
  results$ncoexisting[i] <- sum(xstar > 0)
}
pl <- ggplot(data = results) + 
  aes(x = ncoexisting) + 
  geom_bar() + 
  scale_x_continuous("number of species coexisting", breaks = 0:10)
show(pl)
# add binomial distribution
tbinom <- data.frame(ncoexisting = 0:n, 
                 expectation = nsim * dbinom(0:n, n, 0.5))
pl <- pl + geom_point(data = tbinom, aes(x = ncoexisting, y = expectation))
show(pl)
```

We find again that the number of coexisting species follows the binomial distribution with parameters $n$ and $1/2$ (i.e., exactly the same result found for non interacting species). To prove this fact, let's begin with showing that the probability of having all species coexisting is $1 / 2^n$. For this, we need $x^\star = -A^{-1}r$ to be feasible. For each possible $A$ (invertible) and $r$, we have that the vector $- A^{-1}r$ will display a certain pattern of signs. Proving that the probability that all $n$ species coexist is $1 / 2^n$ amounts to proving that all sign patterns are equally probable. To this end, define the matrix $D_k = (-1)^{\delta_{ik}} \delta_{ij}$. This is like the identity matrix, but with element $k$ of the diagonal set to $-1$ instead of $1$. We have that:

$$
(D_k A D_k) D_k x^\star = -D_k r
$$

That is, by setting the $k^\text{th}$ element of $D_k$ to -1, we flip the sign of the $k^\text{th}$ component of $x^\star$. For example:

```{r proofrz, warning=FALSE}
set.seed(2)
A <- build_LDstable(3)
r <- build_randomr(3)
# the equilibrium is not feasible
solve(A, -r)
# build matrix D_1
D_1 <- diag(c(-1, 1, 1))
# the multiplication changes the sign of a component of the solution
# making the equilibrium feasible
as.numeric(solve(D_1 %*% A %*% D_1, -D_1 %*% r))
```

Because of the symmetry assumption (i.e., we sampled all growth rates and off-diagonal elements from a distribution centered at zero), $(D_k A D_k)$ has the same distribution as $A$, and $D_k r$ the same distribution as $r$. In fact, one can see that $(D_k A D_k)$ is also a similarity transformation, implying that the eigenvalues of $A$ and $(D_k A D_k)$ are the same (i.e., the operation preserves stability):

```{r proofrzsimilarity}
eigen(A)$values
eigen(D_1 %*% A %*% D_1)$values
```

We can repeat the operation several times, connecting each possible starting point to the feasible solution. Therefore, the probability of all species coexisting amounts to the probability of having chosen the "right" sequence of $D_k$, which happens with probability $1 / 2^n$. Notice that the same proof holds when the coefficients of $A$ are sampled in pairs from a bivariate distribution (rather than having symmetric matrices), as long as the distribution is centered at zero, and the matrix is Lyapunov-Diagonally stable (@servan2018coexistence). Using the same argument, one can prove that, under this parametrization, the probability of observing $k$ species coexisting and $n -k$ not being able to invade is exactly $\binom{n}{k} \frac{1}{2^n}$ (@servan2018coexistence).

We can complicate the model substantially, but the results are qualitatively unchanged: the more species ones put in, the more species one finds after dynamics have elapsed. The proportion of species surviving depends on the choice of the distributions for the growth rates and interactions.

## A random zoo in the wild

One might object that the derivations above have no parallel in reality. However, I think that they can be seen as a simplified version of an experiment done routinely by ecologists. Take for example the work of @bittleston2020context. They sampled bacterial communities from ten different pitcher plants, and cultured them in the laboratory on a synthetic medium. They tracked the change of richness in time. This is their Figure 2 (look at panel b):

```{r, echo=FALSE}
url <- "https://media.springernature.com/full/springer-static/image/art%3A10.1038%2Fs41467-020-15169-0/MediaObjects/41467_2020_15169_Fig2_HTML.png"
```
<center><img src="`r url`"></center>

The more the species present at day 3 (i.e., once those with a negative growth rate had disappeared), the more one would find after two months---and the relationship is linear!

## Assembly and saturated equilibrium

Now let's make it more complicated: we assemble an ecological community from the ground up. At each step, we introduce a species at low abundance, starting from an empty community. We then compute the new equilibrium, completing a step of the assembly:

```{r assembly}
assembly_one_step <- function(x, r, A){
  n <- nrow(A)
  invader <- sample(1:n, 1)
  x[invader] <- 0.001 # introduce the invader at low abundance
  present <- x > 0 # these are the species present now
  # compute new equilibrium
  y <- get_final_composition(A[present, present, drop = FALSE], r[present])
  x[present] <- y
  return(x)
}
```

Now we can take a species pool along with their parameters, and try to assemble the system until we can no longer add any more species (i.e., until we reach a saturated equilibrium):

```{r rndzooassembly}
set.seed(7)
n <- 10
A <- build_LDstable(n)
r <- build_randomr(n)
# start with no species
x <- rep(0, n)
# assemble for 40 steps and keep track of richness and composition
ninvasions <- 40
results <- tibble(invasion = 1:ninvasions, 
                  richness = rep(NA, ninvasions),
                  composition = rep(NA, ninvasions))
for (i in 1:ninvasions){
  x <- assembly_one_step(x, r, A) 
  results$richness[i] <- sum(x > 0)
  results$composition[i] <- paste((1:n)[x>0], collapse = "-")
}
knitr::kable(head(results))
pl <- ggplot(results) + aes(x = invasion, y = richness, label = composition) + 
  geom_point() + geom_line() + geom_text(hjust = 0, nudge_y = 0.25, angle = 90)+ 
  ylim(c(0, n * 1.25))
show(pl)
```

Now let's roll back history and assemble again:

```{r alternativehistory}
# start with no species
x <- rep(0, n)
# assemble for 40 steps and keep track of richness
ninvasions <- 40
results <- tibble(invasion = 1:ninvasions, 
                  richness = rep(NA, ninvasions),
                  composition = rep(NA, ninvasions))
for (i in 1:ninvasions){
  x <- assembly_one_step(x, r, A) 
  results$richness[i] <- sum(x > 0)
  results$composition[i] <- paste((1:n)[x>0], collapse = "-")
}
knitr::kable(head(results))
pl <- ggplot(results) + aes(x = invasion, y = richness, label = composition) + 
  geom_point() + geom_line() + geom_text(hjust = 0, nudge_y = 0.25, angle = 90)+ 
  ylim(c(0, n * 1.25))
show(pl)
```

As you can see, despite taking a different assembly history, we reach the same final composition. In fact, this is exactly what we would expect if we were to throw all species in the environment at the same time:

```{r checkfinal}
get_final_composition(A, r)
x
```

As such, given enough time, any assembly history for a symmetric, stable matrix $A$ will eventually reach the final composition represented by the saturated equilibrium (@servan2020tractable).

Interestingly, this is not the case when the matrix is not symmetric. @servan2018coexistence conjectured however that the probability of finding a system whose final composition cannot be assembled one species at a time decreases rapidly with the size of the pool, as long as $A$ is Lyapunov Diagonally stable.

## Network spandrels

What is the network structure of the assembled community vs. that of the initial pool? Can we detect a signature of the forces acting on the community such that some species can persist, while other go extinct? To answer these questions, we start by considering a larger pool of species:

```{r beforeafter}
set.seed(1)
# initial pool
A <- build_LDstable(200)
r <- build_randomr(200)
# final composition
xstar <- get_final_composition(A, r)
A_pruned <- A[xstar > 0, xstar > 0]
r_pruned <- r[xstar > 0]
```

Are the properties of matrix $A$ different from those of the pruned version $\tilde{A}$? We probe this in two ways: first, we build a graph with the strongest interactions, and plot it.

```{r plotnetworksrz}
plot_graph_strong <- function(B, quantile = 0.75){
  Bstrong <- abs(B)
  diag(Bstrong) <- 0
  Bstrong[Bstrong < quantile(Bstrong, quantile)] <- 0
  gr <- graph_from_adjacency_matrix((Bstrong > 0) * 1, mode = "undirected")
  plot(gr, vertex.size=10, vertex.label=NA, layout=layout_with_fr)
}

plot_graph_strong(A)
plot_graph_strong(A_pruned)
```

There seems to be no special structure. A more powerful way to show the same is to plot the eigenvalues of $A$ and $\tilde{A}$. For a symmetric matrix with off-diagonal elements centered at zero, the eigenvalues should follow Wigner's semicircle law:

```{r ploteigenrz}
plot_eigen <- function(B){
  evals <- data.frame(lambda = eigen(B, only.values = TRUE, symmetric = TRUE)$values)
  ggplot(data = evals) + 
    aes(x = lambda) + 
    geom_histogram(bins = as.integer(nrow(evals) / 10), colour = "black")
}
plot_eigen(A)
plot_eigen(A_pruned)
```

As such, the matrix of interactions before/after dynamics seem to have the same properties. However, as shown in @servan2018coexistence, the distribution of the growth rates changes in a non-trivial way:

```{r plotrbeforeafter, warning=FALSE, message=FALSE}
toplot <- data.frame(r = r, type = "before")
toplot <- rbind(toplot, data.frame(r = r_pruned, type = "after"))
ggplot(toplot) + aes(x = r, fill = type) + geom_histogram(position = "dodge")
```

But what if species were to be related to each other? For example, suppose $r_i = 1$ for all species, and build a matrix in which the interactions of species $i + 1$ are obtained by mutating slightly those of species $i$:

```{r similarrz}
n <- 200
r <- rep(1, 200)
A <- matrix(0, n, n)
# set first species
A[1, 1] <- -1
# now each species is obtained by mutating the previous one
for (i in 2:n){
  ai <- A[i -1,] * (1 - 0.05 * runif(n))
  A[i, ] <- A[i, ] + ai
  A[, i] <- A[, i] + ai
  A[i, i] <- -1
}
# make LD-stable
l1 <- max(eigen(A, only.values = TRUE, symmetric = TRUE)$values)
if (l1 > 0) diag(A) <- diag(A) - l1 * 1.01
```

Now each species is similar to the previous one:
```{r similarimage}
image(A[1:n, n:1])
```

Perform the pruning, and plot networks and eigenvalues:

```{r structruredspandrels, warning=FALSE}
xstar <- get_final_composition(A, r)
A_pruned <- A[xstar > 0, xstar > 0]
r_pruned <- r[xstar > 0]

plot_graph_strong(A)
plot_graph_strong(A_pruned)
plot_eigen(A)
plot_eigen(A_pruned)
```

Meaning that if our matrix $A$ is structured, we will recover a structured matrix after pruning, while if $A$ is unstructured, we will recover an unstructured matrix. @maynard2018network showed that a well-defined network structure could be a "network spandrel" (cfr. @gould1979spandrels and @sole2006network) arising from the way new species are introduced, rather than a "signature of stability". 

